---
title: "Performance and Scale: Making R Work Efficiently"
---

```{=latex}
\renewcommand{\thefigure}{24.\arabic{figure}}

\setcounter{figure}{0}

\renewcommand{\thetable}{24.\arabic{table}}

\setcounter{table}{0}
```

```{r}
#| label: setup-chapter-24
#| include: false

suppressPackageStartupMessages({
  library(knitr)
})

knitr::opts_chunk$set(
  message = FALSE,
  warning = FALSE,
  comment = "#>",
  fig.align = "center"
)

set.seed(24)
```

```{r}
#| label: setup-chapter-24-data
#| include: false

if (file.exists("data/processed/df_analysis.rds")) {
  df_ch24 <- readRDS("data/processed/df_analysis.rds")
} else if (file.exists("data/processed/df_analysis.csv")) {
  df_ch24 <- read.csv(
    "data/processed/df_analysis.csv",
    stringsAsFactors = FALSE
  )
} else {
  df_ch24 <- data.frame(
    learner_id = sprintf("L%03d", 1:120),
    country = sample(
      c("Nigeria", "Kenya", "Ghana", "South Africa"),
      120,
      replace = TRUE
    ),
    background = sample(
      c("Research", "Marketing", "Education", "Health"),
      120,
      replace = TRUE
    ),
    study_hours = round(runif(120, 1, 12), 1),
    lessons_completed = sample(4:24, 120, replace = TRUE),
    confidence_score = sample(45:95, 120, replace = TRUE),
    project_score = sample(40:98, 120, replace = TRUE),
    completion_status = sample(
      c("Completed", "In progress", "Paused"),
      120,
      replace = TRUE,
      prob = c(0.55, 0.35, 0.10)
    ),
    stringsAsFactors = FALSE
  )
}
```

## Opening Scene: The Script Works, but It No Longer Feels Light

Leo has a script that works.

That should feel like a victory, and for a while it does. The script imports the learner data, cleans the fields, prepares a few summaries, creates figures, and helps rebuild the report. After the work in [Chapter 23](23-reliability-and-automation.qmd), the order is clearer. The project no longer depends only on what Leo remembers. It has scripts, checks, outputs, and a route.

Then the project grows.

The raw file gets larger. The report now has more tables. The plots take longer to rebuild. A transformation that felt instant with a small dataset now pauses long enough for Leo to wonder whether R has frozen. A colleague asks for the same report by country. Another asks for it by learning background. Someone wants a new version after fresh data arrives.

Nothing is obviously broken. That is what makes the problem more subtle.

The code is correct enough to run. The workflow is reliable enough to repeat. But the work has started to feel heavy.

This is the moment when many learners make the wrong move. They search for "fast R code" and begin replacing clear code with clever fragments they barely understand. The script becomes harder to read. The output may change. The real bottleneck may remain untouched.

Leo needs a better question.

Not, "How do I make everything faster?"

The better question is:

> What part of this workflow is slow enough, large enough, or repeated enough to deserve attention?

That question is the doorway into performance and scale.

## The Lightbulb Idea

::: {.callout-tip}
## The lightbulb idea

Performance is not about making every line faster. It is about knowing where time, memory, and repetition actually matter.
:::

Performance is a professional judgement problem before it is a coding problem. A slow line of code may not matter if it runs once and finishes in less than a second. A moderately slow operation may matter greatly if it runs inside a loop, rebuilds every report, or repeats across many files.

Scale also has a practical meaning. It does not always mean millions of rows or a server. Scale begins when a workflow that was comfortable at one size becomes uncomfortable at the next size. That discomfort may appear as waiting time, memory pressure, repeated manual work, long render times, or fragile code that becomes harder to modify as the project expands.

Chapter 24 therefore does not treat speed as a contest. It treats efficiency as part of professional analytical control.

## Why Performance Belongs Near the End of the Book

Performance appears late in this book for a reason.

Earlier chapters gave Leo the basic language of R: values, types, structures, syntax, logic, functions, and debugging. Part 3 then moved him into data workflows: importing, cleaning, transforming, and summarising. Part 4 helped him turn analysis into visual and narrative communication. Part 5 has moved from reports into systems: reproducible documents, data connections, collaboration, reusable tools, reliability, and automation.

Only now does performance become safe to discuss.

Before Leo understands correctness, performance advice can be dangerous. It may teach him to make wrong code faster. Before Leo understands reproducibility, performance advice can also be distracting. It may encourage quick local fixes that cannot be rebuilt later. Before Leo understands workflow order, performance advice may target a small line while the real delay comes from repeated rendering, repeated imports, unnecessary file writing, or hidden manual steps.

@fig-24-performance-position shows the position of this chapter in Part 5. The movement is not from slow code to fast code alone. It is from communicable work, to traceable work, to reusable work, to repeatable work, to efficient work, and finally to integrated work.

```{r}
#| label: fig-24-performance-position
#| fig-cap: "The Position of Performance and Scale within Part 5."
#| fig-alt: "A simple horizontal flow showing Part 5 moving from reports to connections, collaboration, packages, reliability, performance, and final integration."

plot.new()
plot.window(xlim = c(0, 7), ylim = c(0, 2))
steps <- c(
  "Reports",
  "Connections",
  "Collaboration",
  "Reusable\ntools",
  "Reliability",
  "Performance",
  "Integration"
)
x <- seq(0.7, 6.3, length.out = length(steps))
y <- rep(1, length(steps))

points(x, y, pch = 21, cex = 2.8)
text(x, y, labels = 1:length(steps), cex = 0.8)
text(x, y - 0.45, labels = steps, cex = 0.78)
arrows(
  x[-length(x)] + 0.25,
  y[-length(y)],
  x[-1] - 0.25,
  y[-1],
  length = 0.08
)
```

This is why the chapter sits between [Chapter 23](23-reliability-and-automation.qmd) and [Chapter 25](25-final-integration-project.qmd). Chapter 23 asked whether the project could run again from its own files. Chapter 24 asks whether that repeatable project can remain efficient as the work grows. Chapter 25 will ask Leo to connect everything into one end-to-end analytical system.

## Correctness Before Speed

The first rule of performance is almost disappointing:

> Make the code correct before trying to make it fast.

A fast mistake is still a mistake. A beautifully optimised pipeline that silently changes the meaning of a variable is worse than a slow pipeline that preserves the evidence. Performance work must therefore begin with a known result.

Leo should first ask:

- Does the current code produce the correct output?
- Are the variables interpreted correctly?
- Are missing values handled deliberately?
- Are grouping and filtering decisions documented?
- Can the current result be reproduced after restarting R?
- Is there a simple test, summary, or validation check that protects the result?

Only after these questions are settled should he ask whether the code is too slow, too memory-heavy, or too repetitive.

::: {.callout-warning}
## Do not optimise uncertainty

If Leo does not yet trust the output, he should not begin by changing the code for speed. First make the result trustworthy. Then make the trustworthy result faster if the workflow genuinely needs it.
:::

@tbl-24-correctness-before-speed separates three questions that beginners often collapse into one. Keeping them apart protects the analysis.

```{r}
#| label: tbl-24-correctness-before-speed
#| tbl-cap: "Three Different Questions That Should Not Be Collapsed into One."

correctness_speed <- data.frame(
  Question = c(
    "Is the result correct?",
    "Is the workflow repeatable?",
    "Is the workflow efficient enough?"
  ),
  MainConcern = c(
    "Meaning, evidence, assumptions, and output validity",
    "Inputs, scripts, order, environment, and rebuildability",
    "Time, memory, repetition, responsiveness, and scale"
  ),
  TypicalRisk = c(
    "Making a wrong result look polished",
    "Creating a result that cannot be recreated",
    "Optimising a small detail while the real bottleneck remains"
  ),
  ProfessionalHabit = c(
    "Check outputs before changing strategy",
    "Run from source in a clean session",
    "Measure the bottleneck before rewriting code"
  ),
  check.names = FALSE
)

knitr::kable(correctness_speed)
```

The distinction in @tbl-24-correctness-before-speed is not theoretical. It changes behaviour. If Leo has a suspicious result, he returns to debugging and validation. If Leo cannot rebuild the project, he returns to reliability and automation. If the result is correct and the workflow is repeatable, but the work is now slow or heavy, he is ready for performance thinking.

## What Scale Means in Practical R Work

Scale is often described as a matter of data size. That is only partly true.

A dataset with many rows can create scale problems. But so can many files, many repeated reports, many plots, many joins, many groups, many package dependencies, or many people working on the same project. Scale appears when a strategy that worked comfortably in a small situation begins to strain under a larger situation.

For Leo, scale may show up in six practical ways:

1. The dataset grows.
2. The number of variables grows.
3. The same analysis must be repeated across groups.
4. Reports must be regenerated frequently.
5. Intermediate objects become large.
6. The project becomes hard to inspect, rerun, or share.

@tbl-24-scale-forms gives Leo a first map. It prevents him from thinking that scale always means "big data" in the dramatic sense.

```{r}
#| label: tbl-24-scale-forms
#| tbl-cap: "Practical Forms of Scale in an R Project."

scale_forms <- data.frame(
  FormOfScale = c(
    "More rows",
    "More columns",
    "More groups",
    "More files",
    "More outputs",
    "More collaborators",
    "More frequent rebuilds"
  ),
  WhatLeoNotices = c(
    "Imports, joins, filters, and summaries take longer",
    "Objects become harder to inspect and carry unnecessary fields",
    "The same logic is repeated across countries, cohorts, or categories",
    "Manual import and cleaning steps become fragile",
    "Tables, plots, and reports take longer to regenerate",
    "Changes become harder to track and explain",
    "Small inefficiencies are repeated many times"
  ),
  BetterQuestion = c(
    "Can I reduce, summarise, or process only what I need?",
    "Can I select relevant columns earlier?",
    "Can I write one reusable pattern instead of copying code?",
    "Can I create a controlled import route?",
    "Can I cache or rebuild outputs deliberately?",
    "Can version control and clear structure reduce confusion?",
    "Can automation avoid repeated manual execution?"
  ),
  check.names = FALSE
)

knitr::kable(scale_forms)
```

The practical lesson from @tbl-24-scale-forms is that scale can be computational, organisational, or both. A workflow can be slow because it performs too much computation. It can also be slow because Leo has to remember too many manual steps.

Performance is therefore not only about speed. It is about reducing unnecessary friction.

## A Diagnostic Map of Common Bottlenecks

A bottleneck is the part of a workflow that limits the whole workflow. If Leo wants to improve performance, he should first find the bottleneck. Guessing is unreliable because the slowest part is not always the most visible part.

A plot may look complex, but the delay may come from reading the same raw file repeatedly. A model may look heavy, but the real cost may come from creating several large intermediate objects. A pipeline may look elegant, but it may carry twenty columns that the final report never uses. A report may feel slow, not because any one chunk is terrible, but because the same expensive work is repeated in several chunks.

@tbl-24-performance-diagnosis gives Leo a practical diagnostic map.

```{r}
#| label: tbl-24-performance-diagnosis
#| tbl-cap: "A Diagnostic Map for Common Performance Bottlenecks."

diagnosis_map <- data.frame(
  Symptom = c(
    "The report takes a long time to render",
    "A transformation pauses for a long time",
    "R becomes sluggish or crashes",
    "The same code appears in many places",
    "Import is slow every time the project runs",
    "A grouped analysis is difficult to extend",
    "Code is faster but harder to trust"
  ),
  LikelyBottleneck = c(
    "Repeated expensive chunks or unnecessary recomputation",
    "Inefficient row-by-row logic, joins, or grouping",
    "Large objects, unnecessary copies, or too many intermediate objects",
    "Manual repetition instead of a function or workflow pattern",
    "Reading large files repeatedly or reading more columns than needed",
    "Copy-paste structure instead of a reusable grouped pattern",
    "Optimisation was done before validation"
  ),
  FirstResponse = c(
    "Identify which chunk is expensive before changing the whole chapter",
    "Compare a clear vectorised or grouped approach with the current approach",
    "Inspect object size and remove unused objects when safe",
    "Name the repeated action as a function or controlled script step",
    "Import once into a processed format when the workflow allows it",
    "Write one pattern that works across groups",
    "Return to correctness checks before keeping the faster version"
  ),
  check.names = FALSE
)

knitr::kable(diagnosis_map)
```

The purpose of @tbl-24-performance-diagnosis is not to make Leo anxious. It gives him a way to slow down his diagnosis. A vague complaint such as "R is slow" becomes a sharper question: "Which part is slow, and why?"

## Timing Code Without Becoming Obsessed with Timing

R includes simple timing tools. The most accessible is `system.time()`. It measures how long an expression takes to run. This is enough for early diagnosis.

@lst-24-simple-timing-template gives a small pattern Leo can reuse when he wants to compare two versions of the same operation. The listing is not evaluated because it is a template for his own project.

```{r}
#| lst-label: lst-24-simple-timing-template
#| eval: false
#| lst-cap: "A Simple Timing Template for Comparing Two Equivalent Approaches."

# First, confirm that both approaches produce the same result.

result_a <- approach_a(data)
result_b <- approach_b(data)

stopifnot(identical(result_a, result_b))

# Then compare timing.

system.time({
  result_a <- approach_a(data)
})

system.time({
  result_b <- approach_b(data)
})
```

The important line in @lst-24-simple-timing-template is not `system.time()`. The important line is `stopifnot(identical(result_a, result_b))`. If the two approaches do not produce the same result, Leo is not comparing performance. He is comparing two different analyses.

The next example is deliberately small. It is not a universal benchmark. It simply shows why vectorised thinking often matters. @tbl-24-vectorisation-timing compares a loop that works one element at a time with a vectorised expression that asks R to operate on the whole vector.

```{r}
#| label: tbl-24-vectorisation-timing
#| tbl-cap: "A Small Machine-Dependent Timing Illustration of Loop-Based and Vectorised Code."

x <- runif(200000)

loop_time <- system.time({
  loop_result <- numeric(length(x))
  for (i in seq_along(x)) {
    loop_result[i] <- sqrt(x[i])
  }
})["elapsed"]

vector_time <- system.time({
  vector_result <- sqrt(x)
})["elapsed"]

same_result <- isTRUE(all.equal(loop_result, vector_result))

timing_table <- data.frame(
  Approach = c("Loop over each element", "Vectorised expression"),
  CodeShape = c(
    "Create an output vector, then fill it one value at a time",
    "Apply `sqrt()` directly to the full vector"
  ),
  ElapsedSeconds = round(c(loop_time, vector_time), 4),
  SameResult = c(same_result, same_result),
  check.names = FALSE
)

knitr::kable(timing_table)
```

The numbers in @tbl-24-vectorisation-timing will vary by machine. That is expected. Leo should read the table as a diagnostic illustration, not as a fixed promise about speed. The general lesson is more durable: when R already knows how to operate on a vector, Leo should usually let it do so.

::: {.callout-caution}
## Timing is not truth by itself

One timing run can be noisy. A small example can behave differently from a real project. Timing is useful for diagnosis, but it should be interpreted with judgement, not worshipped.
:::

For more advanced benchmarking, packages such as `bench` can be useful. This chapter does not require them. The goal here is to build the habit of measuring before rewriting.

## Vectorisation and the Danger of Row-by-Row Thinking

Many learners arrive in R with spreadsheet habits. They imagine data as rows that must be handled one at a time. That mental model is understandable, but it can become inefficient in R.

R is built around vectors. Many R functions expect to receive a whole vector and return a whole vector. This is why expressions such as `sqrt(x)`, `log(income)`, `study_hours * 4`, and `confidence_score >= 70` work without an explicit loop.

Vectorisation is not magic. It simply means Leo expresses the operation in a way that matches how R is designed to work.

Consider a simple learner classification. Leo wants to identify learners whose study hours and project scores indicate strong progress. In row-by-row thinking, he may imagine checking each learner one at a time. In vectorised thinking, he creates the condition across the full columns.

```{r}
#| label: ch24-vectorised-classification

if (all(c("study_hours", "project_score") %in% names(df_ch24))) {
  study_hours_num <- suppressWarnings(as.numeric(df_ch24$study_hours))
  project_score_num <- suppressWarnings(as.numeric(df_ch24$project_score))

  df_ch24$strong_progress <- 
    study_hours_num >= median(study_hours_num, na.rm = TRUE) &
    project_score_num >= median(project_score_num, na.rm = TRUE)
}
```

The code above does not produce a table or figure, so it does not need a caption. It teaches a concept and runs safely. Leo is not asking R to visit each row manually in his code. He is asking R to evaluate a full condition across the relevant columns.

@tbl-24-vectorisation-map makes the shift more explicit.

```{r}
#| label: tbl-24-vectorisation-map
#| tbl-cap: "Moving from Row-by-Row Thinking to Vectorised R Thinking."

vectorisation_map <- data.frame(
  BeginnerThought = c(
    "Check the first learner, then the second, then the third",
    "Add a value to each score one row at a time",
    "Create a category by manually visiting each record",
    "Summarise each group through copied filters"
  ),
  RThought = c(
    "Evaluate the condition across the full vector",
    "Use arithmetic directly on the full numeric vector",
    "Use a vectorised condition such as `ifelse()` or `case_when()` where available",
    "Group once and summarise by the grouping variable"
  ),
  WhyItScalesBetter = c(
    "The code describes the whole operation, not each manual step",
    "R uses vector operations instead of repeated user-level instructions",
    "The rule is easier to inspect and less likely to drift across rows",
    "The same logic works when more groups appear"
  ),
  check.names = FALSE
)

knitr::kable(vectorisation_map)
```

The mental shift in @tbl-24-vectorisation-map also improves readability. A clear vectorised expression often says what Leo means more directly than a long loop.

This does not mean loops are forbidden. Loops are useful when each step depends on the previous step, when a task is genuinely iterative, or when Leo is coordinating repeated operations carefully. The mistake is not using a loop. The mistake is using row-by-row code when a clearer vectorised expression already exists.

## Runtime Growth: Why Small Examples Can Mislead

A script can feel fast with 100 rows and slow with 100,000 rows. That does not mean R has become irrational. It often means the cost of an operation grows with the size of the input.

@fig-24-runtime-growth gives a conceptual picture of runtime growth. The curves are not measured from a specific computer. They are a teaching illustration that helps Leo understand why some strategies become uncomfortable faster than others.

```{r}
#| label: fig-24-runtime-growth
#| fig-cap: "A Conceptual Illustration of How Different Runtime Patterns Grow as Work Size Increases."
#| fig-alt: "A line chart showing three conceptual runtime patterns. Linear growth rises slowly, n log n growth rises moderately, and quadratic growth rises sharply as work size increases."

n <- seq(1, 100, by = 1)
linear <- n / max(n)
nlogn <- (n * log(n + 1)) / max(n * log(n + 1))
quadratic <- (n^2) / max(n^2)

matplot(
  n,
  cbind(linear, nlogn, quadratic),
  type = "l",
  lty = c(1, 2, 3),
  lwd = 2,
  xlab = "Work size",
  ylab = "Relative runtime",
  main = "Conceptual Runtime Growth"
)
legend(
  "topleft",
  legend = c("Linear", "n log n", "Quadratic"),
  lty = c(1, 2, 3),
  lwd = 2,
  bty = "n"
)
```

The lesson from @fig-24-runtime-growth is not that Leo must learn formal algorithm analysis before doing data work. The lesson is more practical: a strategy that repeats unnecessary work may look harmless in a small example, then become painful when the input grows.

This is why performance thinking begins with repeated work. A small inefficiency inside a repeated operation can become a large delay.

## Memory, Copying, and Object Size

Time is not the only performance issue. Memory matters too.

In R, objects live in memory during a session. If Leo creates several large copies of the same dataset, memory use can grow quickly. This is not usually a problem with small teaching examples. It can become a problem when a project carries many large intermediate objects, imports unnecessary columns, or repeatedly saves and reloads heavy objects.

A practical first step is to inspect object size. @tbl-24-object-size uses the current analysis dataset when available. If the project data is not present during rendering, the chapter uses a small teaching fallback created in the setup chunk.

```{r}
#| label: tbl-24-object-size
#| tbl-cap: "A Small Object-Size Check Using the Chapter Dataset."

preferred_columns <- intersect(
  c(
    "learner_id",
    "country",
    "background",
    "study_hours",
    "lessons_completed",
    "confidence_score",
    "project_score",
    "completion_status",
    "strong_progress"
  ),
  names(df_ch24)
)

if (length(preferred_columns) == 0) {
  preferred_columns <- names(df_ch24)[seq_len(min(6, ncol(df_ch24)))]
}

df_ch24_subset <- df_ch24[, preferred_columns, drop = FALSE]

object_size_table <- data.frame(
  Object = c("Full chapter dataset", "Selected working columns"),
  Rows = c(nrow(df_ch24), nrow(df_ch24_subset)),
  Columns = c(ncol(df_ch24), ncol(df_ch24_subset)),
  ApproximateMemory = c(
    format(object.size(df_ch24), units = "auto"),
    format(object.size(df_ch24_subset), units = "auto")
  ),
  check.names = FALSE
)

knitr::kable(object_size_table)
```

@tbl-24-object-size is intentionally modest. It does not tell Leo to cut every dataset down immediately. It teaches a habit: inspect what the workflow is carrying. If a report needs five columns, carrying twenty-eight columns through every step may not be harmful at this size, but the habit can become expensive at larger sizes.

::: {.callout-tip}
## A practical memory habit

Select the columns needed for a task when the purpose is clear. Do not remove useful evidence too early, but do not carry unnecessary weight through every later step simply because the full dataset was imported.
:::

The same judgement applies to intermediate objects. Clear names such as `raw_migration`, `df_r_migration`, and `df_analysis` help preserve workflow meaning. But creating many nearly identical objects can make both memory and interpretation harder. Leo should keep intermediate objects when they mark meaningful workflow stages. He should avoid keeping them merely because he is afraid to overwrite anything.

The reliable solution is not reckless overwriting. It is controlled staging: raw data stays raw, processed data is saved deliberately, analysis-ready data is created through code, and temporary objects are removed when they no longer serve a purpose.

## Efficient Pipelines Without Sacrificing Clarity

Efficiency should not destroy readability.

A pipeline that no one understands is not a professional achievement. A faster script that Leo cannot safely modify next month is a liability. The goal is to write code that is clear enough to audit and efficient enough for the work it must do.

This is especially important with `dplyr` pipelines. In the earlier data chapters, pipelines helped Leo express a sequence of data actions. That remains valuable. The question in Chapter 24 is not whether pipelines are good or bad. The question is whether the pipeline is doing unnecessary work.

A clear pipeline usually scales better when it:

- selects only needed columns once the purpose is known
- filters irrelevant rows before expensive summaries when appropriate
- groups deliberately, then ungroups when grouping is no longer needed
- avoids repeating the same transformation in several places
- creates named helper functions for repeated logic
- checks outputs after any major rewrite

@tbl-24-pipeline-discipline summarises the difference between clear inefficient code, clever risky code, and disciplined efficient code.

```{r}
#| label: tbl-24-pipeline-discipline
#| tbl-cap: "Balancing Clarity and Efficiency in Data Pipelines."

pipeline_discipline <- data.frame(
  Pattern = c(
    "Clear but wasteful",
    "Fast but obscure",
    "Disciplined and efficient"
  ),
  WhatItLooksLike = c(
    "The same summaries are recalculated in many chunks",
    "Dense code uses unfamiliar tricks without explanation",
    "The workflow removes unnecessary repetition while keeping intent visible"
  ),
  Risk = c(
    "The project becomes slow as outputs multiply",
    "Future readers cannot safely audit or revise the work",
    "Requires judgement about what should be abstracted or simplified"
  ),
  BetterHabit = c(
    "Save or reuse meaningful intermediate summaries",
    "Explain the strategy or choose a clearer approach",
    "Make the repeated action visible through a function, script, or named object"
  ),
  check.names = FALSE
)

knitr::kable(pipeline_discipline)
```

The middle row in @tbl-24-pipeline-discipline is the trap. Leo should not confuse cleverness with professionalism. Good performance work makes the project easier to run and still possible to understand.

## Larger-Than-Comfortable Data: Practical Options

Most R users do not begin with truly massive data. They begin with data that becomes larger than their current habits can comfortably handle. This may mean a CSV file that takes too long to read, a join that strains memory, or a report that repeats expensive work.

Before reaching for advanced tools, Leo should ask whether a simpler improvement is enough.

Can he filter earlier? Can he select fewer columns? Can he avoid reading the same file repeatedly? Can he save a processed version after cleaning? Can he summarise before plotting? Can he move repeated code into a function from the habits developed in [Chapter 22](22-from-functions-to-packages.qmd)?

If the answer is no, then specialised tools may be appropriate. @tbl-24-tool-options gives a careful map. It does not tell Leo to use every tool. It helps him know when each class of tool becomes relevant.

```{r}
#| label: tbl-24-tool-options
#| tbl-cap: "A Practical Map of R Performance Options as Projects Grow."

tool_options <- data.frame(
  Option = c(
    "Clear base R",
    "Tidyverse workflow",
    "data.table",
    "Database-backed workflow",
    "Arrow-style columnar workflow",
    "Parallel computing",
    "Project redesign"
  ),
  BestWhen = c(
    "The task is small, direct, and easy to express with built-in functions",
    "The priority is readable data manipulation and communicable analysis",
    "Large in-memory tables require fast filtering, grouping, joining, or updating",
    "The data is too large or too centralised to load fully into R",
    "The workflow benefits from columnar storage, partitioned data, or cross-tool access",
    "Independent tasks are large enough to justify worker setup and coordination",
    "The bottleneck is repeated manual or structural work rather than one slow function"
  ),
  Caution = c(
    "Can become hard to read if over-compressed",
    "Can become slow if the same expensive work is repeated without need",
    "Requires learning a different syntax and reference semantics",
    "Requires database knowledge and careful query design",
    "Adds a storage and tooling layer that beginners should not adopt casually",
    "Can increase memory use, complexity, and reproducibility demands",
    "Requires discipline, not a new package"
  ),
  RenderingRoleInThisChapter = c(
    "Safe for evaluated examples",
    "Appropriate where already used in the book setup",
    "Optional demonstration only",
    "Optional demonstration only",
    "Optional demonstration only",
    "Optional demonstration only",
    "Central professional habit"
  ),
  check.names = FALSE
)

knitr::kable(tool_options)
```

The strongest line in @tbl-24-tool-options may be the last one. Sometimes the problem is not that Leo lacks a faster package. The problem is that the workflow is poorly shaped. A poorly shaped workflow can make any package feel slow.

::: {.callout-note}
## Optional tools are not beginner requirements

Packages such as `data.table`, `arrow`, and `duckdb` can be powerful. This chapter mentions them because professionals may encounter them as projects grow. They are not required for Leo to understand performance thinking, and the optional examples below are not evaluated during rendering.
:::

@lst-24-optional-data-table-pattern shows the kind of pattern Leo may later see in `data.table`. It is included for recognition, not as a required dependency.

```{r}
#| lst-label: lst-24-optional-data-table-pattern
#| eval: false
#| lst-cap: "An Optional data.table Pattern for Fast Grouped Summaries."

library(data.table)

learner_dt <- as.data.table(df_analysis)

country_summary <- learner_dt[
  ,
  .(
    learners = .N,
    average_project_score = mean(project_score, na.rm = TRUE),
    average_study_hours = mean(study_hours, na.rm = TRUE)
  ),
  by = country
]
```

@lst-24-optional-database-pattern shows a database-backed pattern. The point is that R does not always need to pull every row into memory before analysis begins.

```{r}
#| lst-label: lst-24-optional-database-pattern
#| eval: false
#| lst-cap: "An Optional Database-Backed Pattern for Working Without Loading All Rows into Memory."

library(DBI)
library(duckdb)
library(dplyr)

con <- dbConnect(duckdb())

learner_tbl <- tbl(con, "df_analysis")

country_summary <- learner_tbl |>
  group_by(country) |>
  summarise(
    learners = n(),
    average_project_score = mean(project_score, na.rm = TRUE),
    .groups = "drop"
  ) |>
  collect()
```

@lst-24-optional-arrow-pattern shows an Arrow-style pattern. Again, this is a recognition example. It should not be treated as a required method for the book.

```{r}
#| lst-label: lst-24-optional-arrow-pattern
#| eval: false
#| lst-cap: "An Optional Arrow-Style Pattern for Columnar Data Workflows."

library(arrow)
library(dplyr)

learner_data <- open_dataset("data/processed/learner_parquet/")

country_summary <- learner_data |>
  select(country, project_score, study_hours) |>
  group_by(country) |>
  summarise(
    average_project_score = mean(project_score, na.rm = TRUE),
    average_study_hours = mean(study_hours, na.rm = TRUE),
    .groups = "drop"
  ) |>
  collect()
```

The three optional listings, @lst-24-optional-data-table-pattern, @lst-24-optional-database-pattern, and @lst-24-optional-arrow-pattern, are deliberately marked `eval: false`. They introduce possibilities without making the chapter depend on additional tools. That is part of render-safe teaching.

## Repetition Tools Are Not Magic Speed Buttons

When Leo begins thinking about performance, he may hear advice such as "use `apply()` instead of loops" or "use `foreach()` to make R faster." These statements contain a useful instinct, but they are too simple.

The deeper issue is not whether a particular function looks more advanced. The deeper issue is whether the work is being expressed at the right level. If Leo repeats the same operation manually, a structured repetition tool may make the code clearer. Sometimes it may also help performance. But no repetition tool is a magic speed button.

The `apply` family is part of base R. It helps Leo express repeated work across vectors, lists, matrices, or margins of a data structure. Its first benefit is structure and clarity. It may avoid clumsy beginner loops, but it is not automatically faster than a well-written loop.

The `purrr::map()` family serves a similar workflow purpose in tidyverse-style projects. It can make repeated actions easier to read and combine. It is useful when the book's earlier function habits begin to meet larger repeated workflows.

`foreach()` is different. It is often used with a parallel backend so that independent tasks can run on multiple workers. That makes it relevant to performance and scale, but only under stricter conditions. The task must be divisible, each piece must be independent, and the overhead of starting and coordinating workers must be worth paying.

@tbl-24-repetition-tools separates these ideas so Leo does not confuse organised repetition with guaranteed speed.

```{r}
#| label: tbl-24-repetition-tools
#| tbl-cap: "A Practical Distinction among Loops, apply, purrr, and foreach."

repetition_tools <- data.frame(
  ToolOrPattern = c(
    "A clear `for` loop",
    "Base R `apply` family",
    "Tidyverse `purrr::map()` family",
    "`foreach()` without a parallel backend",
    "`foreach()` with a parallel backend"
  ),
  BestUse = c(
    "Coordinating repeated work step by step when clarity matters",
    "Applying a function across vectors, lists, matrices, or margins",
    "Applying functions across lists or nested workflow objects in a tidy style",
    "Expressing repeated work with a flexible loop-like syntax",
    "Dividing independent, expensive tasks across several workers"
  ),
  PerformanceMessage = c(
    "Can be perfectly acceptable when written clearly and not doing unnecessary work",
    "Not automatically faster; often better because it expresses repetition cleanly",
    "Usually chosen for clarity and workflow consistency rather than guaranteed speed",
    "May improve structure but does not create parallel execution by itself",
    "Can reduce elapsed time when tasks are independent and large enough to justify overhead"
  ),
  Caution = c(
    "Avoid growing objects inside the loop without need",
    "Do not use it merely to look more advanced",
    "Requires an additional package if not already part of the project setup",
    "Still runs sequentially unless a parallel backend is registered",
    "Requires extra packages, memory, worker setup, and careful reproducibility habits"
  ),
  check.names = FALSE
)

knitr::kable(repetition_tools)
```

The practical lesson from @tbl-24-repetition-tools is simple: Leo should use repetition tools to make repeated work more controlled. He should use parallel tools only when the work is genuinely suitable for parallel execution.

## When One Core Is Not Enough: A Careful Introduction to Parallel Computing

Most ordinary R code runs in one main R process and usually uses one core at a time. This is not a moral failure in R. It is the normal execution model for many interactive analytical tasks.

The qualification matters. Some packages, database engines, and numerical libraries may use more than one core internally. Leo should therefore avoid the careless claim that R can never use more than one core. The safer professional statement is this: Leo should not assume that ordinary R code automatically spreads his work across all available cores.

Parallel computing becomes relevant when Leo explicitly arranges for independent pieces of work to run at the same time. He is setting up workers, giving each worker a portion of the task, and then combining the results.

That sounds attractive, but it has costs. Workers take time to start. Data may need to be copied to workers. Each worker may use memory. Results must be brought back together. Random number generation must be handled carefully if simulation is involved. On a personal computer, using every available core can make the machine unresponsive. It is often better to leave at least one core free for the operating system and other applications.

::: {.callout-warning}
## Parallel is not automatically faster

Parallel computing helps only when the task can be divided safely, each piece is large enough to justify the overhead, and the computer has enough memory to support the workers. Small tasks may become slower when parallelised.
:::

@tbl-24-parallel-judgement gives Leo a disciplined way to decide whether parallel computing belongs in a workflow.

```{r}
#| label: tbl-24-parallel-judgement
#| tbl-cap: "A Practical Judgement Map for Deciding Whether Parallel Computing Is Appropriate."

parallel_judgement <- data.frame(
  Question = c(
    "Can the task be divided into independent pieces?",
    "Does each piece take enough time to justify worker setup?",
    "Will each worker need a large copy of the data?",
    "Can the results be combined safely?",
    "Does the work involve random numbers?",
    "Can the computer spare the cores and memory?"
  ),
  GoodSign = c(
    "Each model, file, group, or simulation can run without depending on the previous one",
    "Each task is substantial enough that coordination cost is small by comparison",
    "Workers can receive small inputs or work from controlled sources",
    "The output has a predictable structure such as rows, lists, or saved files",
    "Seeds and random-number strategy are planned deliberately",
    "At least one core and enough memory remain free for the system"
  ),
  WarningSign = c(
    "Each step depends on the result of the step before it",
    "The task is tiny and the overhead may dominate",
    "Every worker needs a full heavy dataset copied into memory",
    "The combined result may change order, structure, or meaning",
    "Re-running may produce unstable results without careful seed handling",
    "The machine becomes sluggish, unstable, or memory-constrained"
  ),
  check.names = FALSE
)

knitr::kable(parallel_judgement)
```

The questions in @tbl-24-parallel-judgement protect Leo from a common beginner mistake: using more cores before understanding the shape of the work. Parallel computing is strongest when many independent tasks are waiting in a queue. It is weak when the workflow is slow because of repeated imports, unclear code, unnecessary object copies, or a task that cannot be safely divided.

@lst-24-optional-foreach-pattern shows a cautious optional pattern using `foreach` with a parallel backend. The listing is non-evaluated because it depends on optional packages and local machine resources.

```{r}
#| lst-label: lst-24-optional-foreach-pattern
#| eval: false
#| lst-cap: "An Optional foreach Pattern for Independent Parallel Tasks."

library(foreach)
library(doParallel)

available_cores <- parallel::detectCores(logical = TRUE)

if (is.na(available_cores)) {
  available_cores <- 2
}

workers <- max(1, available_cores - 1)

cluster <- parallel::makeCluster(workers)
on.exit(parallel::stopCluster(cluster), add = TRUE)

doParallel::registerDoParallel(cluster)

results <- foreach(
  group_name = unique(df_analysis$country),
  .combine = rbind
) %dopar% {
  group_data <- df_analysis[df_analysis$country == group_name, ]

  data.frame(
    country = group_name,
    learners = nrow(group_data),
    average_project_score = mean(group_data$project_score, na.rm = TRUE)
  )
}
```

@lst-24-optional-foreach-pattern should be read as a pattern, not as a first response to slow code. It is appropriate only when the repeated work is independent, large enough to matter, and worth the extra machinery. For ordinary summaries on modest datasets, a clear grouped summary is usually better.

This is the balanced message Leo needs: parallel computing belongs in his long-term toolbox, but it should not become his first explanation for performance. Before adding workers, he should remove unnecessary repetition, use vectorised operations where they fit, check object size, and measure the actual bottleneck.


## Performance as Workflow Design

By this point, Leo can see that performance is not only inside single lines of code. It is also inside the shape of the workflow.

A workflow that imports the same raw CSV file in ten chunks is inefficient even if each import is correct. A report that recreates the same summary table five times is inefficient even if every table is accurate. A project that forces Leo to rerun a long analysis after changing one sentence in the narrative is inefficient even if the final HTML looks good.

Performance design asks how work should move through the project.

@fig-24-performance-cycle presents a disciplined performance cycle. It begins with correctness and ends with rechecking, not with rewriting everything.

```{r}
#| label: fig-24-performance-cycle
#| fig-cap: "A Disciplined Cycle for Improving Performance Without Losing Trust."
#| fig-alt: "A circular workflow showing six steps: confirm correctness, identify bottleneck, measure enough, change one thing, recheck output, keep readability."

plot.new()
plot.window(xlim = c(0, 10), ylim = c(0, 10))
labels <- c(
  "Confirm\ncorrectness",
  "Identify\nbottleneck",
  "Measure\nenough",
  "Change\none thing",
  "Recheck\noutput",
  "Keep\nreadability"
)
angles <- seq(pi / 2, pi / 2 - 2 * pi + 2 * pi / 6, length.out = 6)
cx <- 5 + 3.2 * cos(angles)
cy <- 5 + 3.2 * sin(angles)

points(cx, cy, pch = 21, cex = 3)
text(cx, cy, labels = seq_along(labels), cex = 0.8)
text(cx, cy - 0.65, labels = labels, cex = 0.75)

for (i in seq_along(cx)) {
  j <- ifelse(i == length(cx), 1, i + 1)
  arrows(cx[i], cy[i], cx[j], cy[j], length = 0.08)
}
text(5, 5, "Performance\njudgement", cex = 0.95)
```

The cycle in @fig-24-performance-cycle protects Leo from two opposite errors. The first is doing nothing while the workflow becomes painful. The second is rewriting everything without knowing what problem he is solving.

This cycle also connects directly to [Chapter 23](23-reliability-and-automation.qmd). A reliable workflow gives Leo something stable to measure. Without stability, performance work becomes guesswork.

## A Small Applied Example Using the Learner Dataset

Now Leo can apply the performance lens to the learner dataset.

Suppose he wants a country-level summary of study behaviour and project performance. The task is ordinary. That is the point. Performance problems often begin in ordinary tasks repeated many times.

The following evaluated chunk uses base R so that the chapter remains render-safe. If the project's `df_analysis` file is available, the setup chunk has already loaded it. Otherwise, the chapter uses a small teaching fallback. @tbl-24-country-summary shows the resulting country-level summary.

```{r}
#| label: tbl-24-country-summary
#| tbl-cap: "A Country-Level Summary Created Once and Reused Deliberately."
required_summary_cols <- c("country", "study_hours", "project_score")
has_required_summary_cols <- all(required_summary_cols %in% names(df_ch24))

if (has_required_summary_cols) {
  summary_data <- data.frame(
    country = df_ch24$country,
    study_hours = suppressWarnings(as.numeric(df_ch24$study_hours)),
    project_score = suppressWarnings(as.numeric(df_ch24$project_score)),
    stringsAsFactors = FALSE
  )

  country_values <- sort(unique(summary_data$country[!is.na(summary_data$country)]))

  learners <- as.integer(table(
    factor(summary_data$country, levels = country_values)
  ))

  average_study_hours <- round(
    tapply(
      summary_data$study_hours,
      summary_data$country,
      mean,
      na.rm = TRUE
    )[country_values],
    2
  )

  average_project_score <- round(
    tapply(
      summary_data$project_score,
      summary_data$country,
      mean,
      na.rm = TRUE
    )[country_values],
    2
  )

  country_summary <- data.frame(
    Country = country_values,
    Learners = learners,
    `Average study hours` = as.numeric(average_study_hours),
    `Average project score` = as.numeric(average_project_score),
    check.names = FALSE
  )
} else {
  country_summary <- data.frame(
    Country = character(),
    Learners = integer(),
    `Average study hours` = numeric(),
    `Average project score` = numeric(),
    check.names = FALSE
  )
}

knitr::kable(country_summary)
```

The point of @tbl-24-country-summary is not that base R is always faster than every alternative. The point is that a meaningful summary should be created once, named clearly, inspected, and reused. If the same summary is needed in several places, Leo should not copy the entire calculation into every report section.

That principle is simple. It has large consequences.

If a workflow needs the same summary for a table, a figure, and a paragraph, the project can create `country_summary` once and use it three times. This improves speed, reduces drift, and makes the logic easier to inspect.

## Reducing Repetition Without Hiding Meaning

Repeated work is one of the easiest performance problems to miss. It often appears as a normal habit: copy a chunk, change a variable, run again. At first, this feels efficient. Later, it becomes expensive.

The correction is not to hide everything inside complicated abstractions. The correction is to name repeated ideas at the right level.

If Leo repeatedly calculates summary statistics for different groups, a small helper function may be appropriate. If he repeatedly renders reports for different audiences, a parameterised Quarto document may be appropriate. If he repeatedly reads and cleans the same raw files, a script route may be appropriate. If he repeatedly uses the same project-specific functions, the package habits from [Chapter 22](22-from-functions-to-packages.qmd) become relevant.

@tbl-24-repetition-responses shows the response that fits each kind of repetition.

```{r}
#| label: tbl-24-repetition-responses
#| tbl-cap: "Matching Different Forms of Repetition to Appropriate Workflow Responses."

repetition_responses <- data.frame(
  RepetitionType = c(
    "Same expression repeated inside one script",
    "Same transformation repeated across scripts",
    "Same report repeated for different groups",
    "Same cleaning route repeated for fresh data",
    "Same project tools reused across projects"
  ),
  BetterResponse = c(
    "Create a named object or small helper function",
    "Move the logic into one script or reusable function",
    "Use parameters or a controlled rendering strategy",
    "Automate the import-clean-prepare route with checks",
    "Organise the tools into package-like structure"
  ),
  PerformanceBenefit = c(
    "Avoids recalculating or rewriting the same logic",
    "Reduces drift and repeated computation",
    "Prevents manual copy-paste report production",
    "Makes repeated updates more efficient and reliable",
    "Turns repeated project knowledge into reusable tools"
  ),
  Continuity = c(
    "Functions from earlier chapters",
    "Workflow discipline from Chapter 23",
    "Quarto reporting from Part 5",
    "Reliability and automation from Chapter 23",
    "Reusable tool organisation from Chapter 22"
  ),
  check.names = FALSE
)

knitr::kable(repetition_responses)
```

The most important word in @tbl-24-repetition-responses is "appropriate." Not every repeated line deserves a function. Not every function deserves a package. Not every report deserves a parameterised system. Performance work should reduce friction without creating unnecessary machinery.

## Common Mistakes Leo Should Avoid

Performance work attracts mistakes because it feels technical. The common errors are not only coding errors. They are judgement errors.

@tbl-24-common-mistakes names the most important ones.

```{r}
#| label: tbl-24-common-mistakes
#| tbl-cap: "Common Beginner Mistakes When Trying to Improve R Performance."

common_mistakes <- data.frame(
  Mistake = c(
    "Optimising before checking correctness",
    "Assuming the visible line is the bottleneck",
    "Replacing clear code with clever code too early",
    "Benchmarking tiny examples and overgeneralising",
    "Loading advanced packages without a real need",
    "Repeating expensive work across chunks",
    "Assuming more cores always means faster work",
    "Keeping every intermediate object forever"
  ),
  WhyItHurts = c(
    "The project may preserve a wrong result more efficiently",
    "Time is spent changing code that does not limit the workflow",
    "The project becomes harder to audit and maintain",
    "Small examples may not represent the actual workflow",
    "Dependency weight increases without solving the problem",
    "Render time grows as the report expands",
    "Parallel overhead and memory pressure may make the workflow slower or less stable",
    "Memory and interpretation both become heavier"
  ),
  BetterHabit = c(
    "Validate first, then optimise",
    "Measure enough to locate the bottleneck",
    "Prefer readable efficiency",
    "Use timing as evidence, not as a universal law",
    "Escalate tools only when the project justifies them",
    "Create once, name clearly, reuse deliberately",
    "Use parallel workers only when the task is independent and worth the overhead",
    "Keep meaningful stages, remove temporary clutter"
  ),
  check.names = FALSE
)

knitr::kable(common_mistakes)
```

The mistakes in @tbl-24-common-mistakes are common because they feel productive. Leo may feel as if he is improving the project when he adds a faster package, rewrites a pipeline, or removes intermediate objects. But the professional test is stricter: did the change solve the real problem without damaging trust?

## Try It Yourself

Choose one script or chapter from your current project and inspect it through the performance lens.

Do not change anything at first. Read it and answer these questions:

1. Which parts are repeated?
2. Which objects are created but never used again?
3. Which chunks recreate the same summary?
4. Which imports happen more than once?
5. Which operations would become uncomfortable if the data had ten times more rows?
6. Which parts must remain clear even if they could be written more compactly?

Then make only one improvement.

Do not rewrite the whole project. Choose one meaningful change: remove one repeated calculation, create one reusable object, select only needed columns for one task, or add one simple timing check around a suspicious section.

After the change, rerun the relevant output and confirm that the result still means the same thing.

## A Small Practice Lab

@lst-24-practice-lab gives Leo a practical scaffold. It is not evaluated because the point is for him to apply it inside his own project, where the real bottlenecks live.

```{r}
#| lst-label: lst-24-practice-lab
#| eval: false
#| lst-cap: "A Practice-Lab Scaffold for Diagnosing Performance and Scale in a Real Project."

# Practice Lab: Make one workflow lighter

# 1. Pick one script, report, or chapter that feels slow or heavy.

# 2. Confirm correctness first:
#    - Does the current output look right?
#    - Are missing values handled deliberately?
#    - Are the important summaries stable after rerunning?

# 3. Identify one suspected bottleneck:
#    - repeated import
#    - repeated summary
#    - unnecessary columns
#    - row-by-row logic
#    - large temporary objects
#    - expensive chunk during render

# 4. Measure enough to avoid guessing:
#    system.time({
#      # place the suspected operation here
#    })

# 5. Make one change:
#    - reuse a named summary
#    - select needed columns
#    - replace row-by-row code with vectorised code
#    - move repeated logic into a function
#    - avoid recalculating the same result in many chunks

# 6. Recheck the output:
#    - Did the result stay the same?
#    - Is the code still readable?
#    - Did the change improve the workflow enough to keep?

# 7. Write a short note:
#    - What was changed?
#    - Why was it changed?
#    - What should a future reader know?
```

The lab in @lst-24-practice-lab teaches a disciplined habit: improve the project where the project asks to be improved. Performance work should respond to evidence.

## Short Recap

Leo began this chapter with a script that worked but no longer felt light. That situation is normal. As data, reports, and repeated requests grow, a correct workflow may begin to feel slow, heavy, or difficult to scale.

He learned that performance is not a race to make every line faster. It is a form of judgement. The first task is to protect correctness. The second is to preserve repeatability. Only then should Leo identify the actual bottleneck and improve the smallest meaningful part of the workflow.

He learned that scale can mean more rows, more columns, more groups, more files, more outputs, more collaborators, or more frequent rebuilds. He also learned that timing tools are useful but limited. They help him diagnose; they do not replace judgement.

He saw why vectorised thinking matters in R, why small examples can mislead, why memory and object size deserve attention, and why efficient pipelines should remain readable. He also learned that `apply`, `purrr`, and `foreach` serve different purposes, and that parallel computing is useful only when independent work is large enough to justify the overhead. Optional tools such as `data.table`, databases, Arrow-style workflows, and parallel backends are possibilities for later growth, not requirements for every project.

Most importantly, Leo learned that performance is part of workflow design. A professional project should not merely run once. It should run again, remain understandable, and avoid unnecessary friction as the work expands.

## Leo's Notebook

- Correctness comes before speed.
- A repeatable workflow is easier to measure than a fragile one.
- "R is slow" is not a diagnosis. Find the bottleneck.
- Performance problems may come from time, memory, repetition, imports, outputs, or workflow design.
- Vectorised code often fits R better than row-by-row thinking.
- A small inefficiency becomes important when it is repeated many times.
- Do not trust a faster rewrite until the output has been checked.
- `apply` and `purrr` organise repetition; they do not guarantee speed by themselves.
- Parallel computing is useful only when independent tasks are large enough to justify worker setup and coordination.
- Optional performance tools are useful only when the problem justifies their cost.
- Readability is part of performance because future work must still be possible.
- The best performance change is often small, specific, measured, and documented.

## Glossary

**Bottleneck**  
The part of a workflow that limits the speed, responsiveness, memory use, or maintainability of the whole project.

**Core**  
A processing unit that can carry out computation. Many computers have multiple cores, but ordinary R code does not automatically distribute every task across all of them.

**Worker**  
A separate R process used in parallel computing to handle one part of a divided task. Workers can reduce elapsed time for suitable tasks, but they also require setup, memory, and coordination.

**Memory**  
The working space R uses to hold objects during a session. Large objects and unnecessary copies can increase memory pressure.

**Object size**  
The approximate amount of memory used by an R object. It can be inspected with `object.size()`.

**Performance**  
The practical behaviour of code or a workflow in relation to time, memory, repetition, and responsiveness.

**Premature optimisation**  
Changing code for speed before confirming correctness, repeatability, or the real bottleneck.

**Parallel computing**  
Running independent pieces of work at the same time through multiple worker processes or cores, usually with extra setup, memory, and coordination costs.

**Repetition tool**  
A function or pattern, such as `lapply()`, `purrr::map()`, or `foreach()`, that helps express repeated work more systematically.

**Scale**  
The point at which a workflow must handle more rows, columns, groups, files, outputs, collaborators, or repeated rebuilds than before.

**Timing**  
Measuring how long code takes to run. In base R, `system.time()` provides a simple first timing tool.

**Vectorisation**  
Writing code so that R operates on whole vectors rather than manually processing one value at a time in user-written row-by-row logic.

**Workflow design**  
The deliberate organisation of inputs, scripts, objects, outputs, checks, and repeated actions so that a project remains clear and workable as it grows.

## Chapter Outcome

By the end of this chapter, Leo can explain why performance should come after correctness and reliability. He can distinguish speed, memory, repetition, and workflow friction. He can use a simple timing pattern to investigate a suspected bottleneck. He understands why vectorised operations often fit R better than row-by-row thinking, and he knows that object size and unnecessary copies can matter as projects grow.

He can also make tool choices more carefully. He knows that base R and tidyverse workflows are often enough, that `apply`-style tools organise repetition rather than automatically reducing time, and that optional tools such as `data.table`, databases, Arrow-style workflows, `foreach`, and parallel backends have specific use cases. New packages and new execution models should not be added merely because they sound advanced.

Most importantly, Leo can now ask a stronger professional question:

> Can this workflow stay clear, correct, repeatable, and efficient as the work grows?

That question completes the final technical discipline before integration.

## What Comes Next

Leo now has the last missing habit before the book's final project.

He can write code. He can debug it. He can import, clean, transform, summarise, visualise, and communicate data. He can build reproducible reports, connect data sources, collaborate through traceable changes, organise reusable tools, make workflows repeatable, and now think about performance and scale.

The remaining task is not another isolated skill. It is integration.

[Chapter 25](25-final-integration-project.qmd) brings the pieces together. It asks Leo to stop seeing the book as separate lessons and start seeing the full analytical route as one connected system: data input, cleaning, transformation, analysis, visualisation, reporting, reliability, automation, performance, and final delivery.

Chapter 24 has asked whether a repeatable workflow can remain efficient as the work grows. Chapter 25 asks whether all the parts can now become one professional analytical product.
